---
title: "Weekly US Gas Prices"
format:
  html:
    toc: true
    toc-location: left
execute:
  echo: true
  warning: false
  message: false
---

## Introduction

Weekly U.S. retail prices for **diesel** and **regular gasoline** from 1995 to 2025 reveal market dynamics, seasonality, and volatility. In this article, we:

1. Trace long-term trends.
2. Uncover seasonal patterns.
3. Quantify week-to-week volatility.
4. Compare diesel vs. gasoline pricing.

---

## Setup

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(scales)
library(viridis)
library(patchwork)

# Global theme
theme_set(
  theme_minimal(base_size = 14) +
    theme(
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      panel.grid.minor      = element_blank()
    )
)

# Create images directory
if (!dir.exists("images")) dir.create("images")
```

## Data Acquisition & Cleaning

```{r data}
# Read and clean data
prices <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-07-01/weekly_gas_prices.csv",
  col_types = cols(date = col_date(), price = col_double())
) %>%
  mutate(
    year  = year(date),
    month = month(date, label = TRUE, abbr = TRUE)
  ) %>%
  filter(
    (fuel == "diesel" & grade == "all") |
    (fuel == "gasoline" & grade == "regular" & formulation == "conventional")
  )
```

## Annual Average Prices

```{r annual-plot}
annual <- prices %>%
  group_by(year, fuel) %>%
  summarize(avg_price = mean(price, na.rm = TRUE), .groups = "drop")

p1 <- ggplot(annual, aes(x = year, y = avg_price, color = fuel)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_color_viridis_d(option = "D", end = .8) +
  scale_x_continuous(breaks = seq(1995, 2025, by = 5)) +
  labs(
    title   = "Average Annual US Gas Prices (1995–2025)",
    x       = "Year",
    y       = "Average Price (USD)",
    color   = "Fuel Type",
    caption = "Data: TidyTuesday · Analysis: Joseph Nyajuoga"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1
```

## Seasonal Patterns

```{r monthly-plot}
p2 <- ggplot(prices, aes(x = month, y = price, fill = fuel)) +
  geom_boxplot(alpha = 0.8, outlier.shape = NA) +
  scale_fill_viridis_d(option = "C") +
  labs(
    title = "Monthly Distribution of Gas Prices",
    x     = "Month",
    y     = "Price (USD)",
    fill  = "Fuel Type"
  ) +
  theme(legend.position = "top")

p2
```

## Volatility: Week-to-Week Change

```{r weekly-change-plot}
diffs <- prices %>%
  arrange(date) %>%
  group_by(fuel) %>%
  mutate(delta = price - lag(price)) %>%
  drop_na()

p3 <- ggplot(diffs, aes(x = date, y = delta, color = fuel)) +
  geom_hline(yintercept = 0, color = "gray70") +
  geom_line(alpha = 0.6) +
  scale_color_viridis_d(option = "A", end = .7) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Weekly Change in Gas Prices",
    x     = "Date",
    y     = "Δ Price (USD)",
    color = "Fuel Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p3
```

## Price Relationship: Diesel vs. Gasoline

```{r scatter-plot}
wide <- prices %>%
  select(date, fuel, price) %>%
  pivot_wider(names_from = fuel, values_from = price)

p4 <- ggplot(wide, aes(x = gasoline, y = diesel, color = date)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  scale_color_viridis_c(option = "B") +
  labs(
    title = "Diesel vs. Regular Gasoline Prices",
    x     = "Gasoline (USD)",
    y     = "Diesel (USD)"
  ) +
  theme(legend.position = "none")

p4
```

## Combined Overview

```{r combined-plot, fig-width=12, fig-height=8, fig-format="svg", out-width="100%"}
combined <- (p1 + p2 + p3 + p4) +
  plot_layout(ncol = 2, guides = "collect", tag_level = "new") &
  theme(plot.margin = margin(10,10,10,10), plot.tag = element_text(face = "bold"))

final_plot <- combined +
  plot_annotation(
    title      = "Weekly US Gas Prices: Trends, Seasonality & Volatility",
    subtitle   = "Diesel vs. Regular Gasoline, 1995–2025",
    caption    = "Data: TidyTuesday · Analysis: Joseph Nyajuoga",
    tag_levels = "A"
  ) &
  theme(
    plot.title    = element_text(size = 18, face = "bold"),
    plot.subtitle = element_text(size = 14),
    plot.caption  = element_text(size = 10)
  )

final_plot
```

## Save High-Resolution Files

```{r save-plots, include=FALSE}
ggsave("images/annual_avg_prices.png",      plot = p1, width = 9,  height = 5, dpi = 600)
ggsave("images/monthly_distribution.png",   plot = p2, width = 8,  height = 5, dpi = 600)
ggsave("images/weekly_change.png",          plot = p3, width = 9,  height = 5, dpi = 600)
ggsave("images/diesel_vs_gasoline.png",     plot = p4, width = 7,  height = 5, dpi = 600)
ggsave("images/final_overview.svg",         plot = final_plot, width = 12, height = 8, device = "svg")
```
