---
title: "Weekly US Gas Prices"
author: "Joseph Nyajuoga"
date: today
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 2
    code-fold: true
    code-summary: "Show code"
execute:
  warning: false
  message: false
brand: _brand.yml
---

## Introduction

Weekly U.S. retail prices for **diesel** and **regular gasoline** span from 1995 to 2025, covering three decades of energy market history. This analysis explores four questions:

1.  How have prices trended over the long run, and what drove the largest swings?
2.  Is there a consistent seasonal pattern within each year?
3.  How volatile are prices week to week, and when were the most turbulent periods?
4.  How closely do diesel and gasoline prices track each other?

The data come from the U.S. Energy Information Administration (EIA) via the TidyTuesday project (2025-07-01 release). Prices are in nominal USD per gallon.

------------------------------------------------------------------------

## Data

```{r setup}
#| include: false
library(tidyverse)
library(lubridate)
library(scales)
library(viridis)
library(patchwork)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title.position   = "plot",
      plot.caption.position = "plot",
      panel.grid.minor      = element_blank(),
      plot.title            = element_text(face = "bold")
    )
)

if (!dir.exists("images")) dir.create("images")
```

```{r data}
prices <- read_csv(
  "https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-07-01/weekly_gas_prices.csv",
  col_types = cols(date = col_date(), price = col_double())
) |>
  mutate(
    year  = year(date),
    month = month(date, label = TRUE, abbr = TRUE)
  ) |>
  filter(
    (fuel == "diesel"   & grade == "all") |
    (fuel == "gasoline" & grade == "regular" & formulation == "conventional")
  )
```

The dataset contains `r nrow(prices)` weekly observations across both fuel types, spanning `r min(prices$date)` to `r max(prices$date)`.

```{r summary-table}
prices |>
  group_by(fuel) |>
  summarise(
    Weeks       = n(),
    `Min (USD)` = min(price,  na.rm = TRUE),
    `Mean (USD)`= mean(price, na.rm = TRUE),
    `Max (USD)` = max(price,  na.rm = TRUE),
    `SD (USD)`  = sd(price,   na.rm = TRUE)
  ) |>
  mutate(across(where(is.double), \(x) round(x, 2))) |>
  rename(Fuel = fuel) |>
  knitr::kable(caption = "Summary statistics for weekly retail prices (1995–2025)")
```

------------------------------------------------------------------------

## Long-Run Trends

```{r annual-plot}
annual <- prices |>
  group_by(year, fuel) |>
  summarise(avg_price = mean(price, na.rm = TRUE), .groups = "drop")

events <- tibble(
  year  = c(2008, 2012, 2020, 2022),
  label = c("2008\nCrisis", "Post-crisis\npeak", "COVID-19\ndrop", "Ukraine\nwar spike"),
  y     = c(3.3, 3.9, 1.0, 4.8)
)

p1 <- ggplot(annual, aes(x = year, y = avg_price, color = fuel)) +
  geom_vline(xintercept = c(2008, 2012, 2020, 2022),
             linetype = "dashed", color = "grey80", linewidth = 0.5) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  geom_text(
    data = events,
    aes(x = year, y = y, label = label),
    inherit.aes = FALSE,
    size = 2.8, color = "grey40", lineheight = 0.9, hjust = 1.1
  ) +
  scale_color_viridis_d(option = "D", end = .8,
                        labels = c("Diesel", "Gasoline")) +
  scale_x_continuous(breaks = seq(1995, 2025, by = 5)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title   = "Annual Average US Retail Fuel Prices (1995–2025)",
    x       = NULL,
    y       = "Average Price (USD/gallon)",
    color   = NULL,
    caption = "Source: EIA via TidyTuesday · Analysis: Joseph Nyajuoga"
  ) +
  theme(
    legend.position   = "top",
    axis.text.x       = element_text(angle = 45, hjust = 1)
  )

p1
```

::: callout-note
Prices roughly doubled between 1995 and their 2012 post-crisis peak, driven by growing global demand and constrained supply. Two sharp reversals stand out: the 2008 financial crisis (demand destruction) and the 2020 COVID-19 lockdowns (the steepest single-year drop in the dataset). The 2022 Russia-Ukraine war sent prices to their highest nominal level on record before the subsequent retreat.
:::

------------------------------------------------------------------------

## Seasonal Patterns

```{r monthly-plot}
p2 <- ggplot(prices, aes(x = month, y = price, fill = fuel)) +
  geom_boxplot(alpha = 0.75, outlier.shape = 21, outlier.size = 1,
               outlier.alpha = 0.4) +
  scale_fill_viridis_d(option = "C", labels = c("Diesel", "Gasoline")) +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title   = "Monthly Price Distribution Across All Years",
    x       = NULL,
    y       = "Price (USD/gallon)",
    fill    = NULL,
    caption = "Source: EIA via TidyTuesday · Analysis: Joseph Nyajuoga"
  ) +
  theme(legend.position = "top")

p2
```

::: callout-note
Gasoline prices show a mild summer premium (May-August), reflecting the U.S. summer driving season and the switch to more expensive summer-blend formulations. Diesel exhibits the opposite tendency: prices are often highest in late autumn and winter when heating oil demand competes with trucking for the same distillate supply. The wide interquartile ranges reflect price-level differences across years more than within-year seasonality.
:::

------------------------------------------------------------------------

## Week-to-Week Volatility

```{r weekly-change-plot}
diffs <- prices |>
  arrange(date) |>
  group_by(fuel) |>
  mutate(delta = price - lag(price)) |>
  drop_na()

extreme_weeks <- diffs |>
  group_by(fuel) |>
  slice_max(abs(delta), n = 1)

p3 <- ggplot(diffs, aes(x = date, y = delta, color = fuel)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.6) +
  geom_line(alpha = 0.55, linewidth = 0.4) +
  geom_point(
    data = extreme_weeks,
    aes(x = date, y = delta),
    size = 3, shape = 21, fill = "white", stroke = 1.2
  ) +
  scale_color_viridis_d(option = "A", end = .7,
                        labels = c("Diesel", "Gasoline")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(labels = dollar_format()) +
  labs(
    title   = "Week-on-Week Change in US Retail Fuel Prices",
    x       = NULL,
    y       = "Price change (USD/gallon)",
    color   = NULL,
    caption = "Source: EIA via TidyTuesday · Analysis: Joseph Nyajuoga"
  ) +
  theme(
    legend.position = "top",
    axis.text.x     = element_text(angle = 45, hjust = 1)
  )

p3
```

```{r volatility-stats}
#| echo: false
diffs |>
  group_by(fuel) |>
  summarise(
    `Mean |Δ| (USD)` = mean(abs(delta), na.rm = TRUE),
    `Max rise (USD)`  = max(delta,       na.rm = TRUE),
    `Max fall (USD)`  = min(delta,       na.rm = TRUE)
  ) |>
  mutate(across(where(is.double), \(x) round(x, 3))) |>
  rename(Fuel = fuel) |>
  knitr::kable(caption = "Week-on-week price-change statistics")
```

::: callout-note
Most weeks see a change of less than five cents per gallon, but the distribution has heavy tails. The largest single-week swings (circled) occur around supply shocks; the 2020 demand collapse and the 2022 war-driven supply shock are both visible as dense clusters of outsized moves. Diesel volatility tends to amplify gasoline moves, reflecting its dual role as both a transport and heating fuel.
:::

------------------------------------------------------------------------

## Diesel vs. Gasoline

```{r scatter-plot}
wide <- prices |>
  select(date, fuel, price) |>
  pivot_wider(names_from = fuel, values_from = price) |>
  drop_na()

cor_val <- cor(wide$diesel, wide$gasoline) |> round(3)

p4 <- ggplot(wide, aes(x = gasoline, y = diesel, color = year(date))) +
  geom_point(alpha = 0.55, size = 1.2) +
  geom_smooth(method = "lm", color = "black", linewidth = 0.8, se = TRUE) +
  scale_color_viridis_c(option = "B", name = "Year") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = dollar_format()) +
  annotate("text", x = 1.4, y = 4.8,
           label = paste0("r = ", cor_val),
           size = 4, hjust = 0, color = "grey30") +
  labs(
    title   = "Diesel vs. Regular Gasoline Prices",
    x       = "Gasoline (USD/gallon)",
    y       = "Diesel (USD/gallon)",
    caption = "Source: EIA via TidyTuesday · Analysis: Joseph Nyajuoga"
  ) +
  theme(legend.position = "right")

p4
```

::: callout-note
Diesel and gasoline prices are highly correlated (r = `r cor_val`), as both are refined from crude oil. However, diesel consistently commands a premium over gasoline — a structural shift that became pronounced after 2005 as developing-world demand for distillates grew and U.S. refiners were already running near capacity. Earlier points (darker colour) cluster in the lower-left, reflecting the lower absolute price level of the 1990s and early 2000s.
:::

------------------------------------------------------------------------

## Combined Overview

```{r combined-plot}
#| echo: false
#| fig-width: 13
#| fig-height: 9
#| out-width: "100%"

# Shared palette — same two colours across all four panels
fp <- c("diesel" = "#440154", "gasoline" = "#5DC863")
fl <- c("diesel" = "Diesel",  "gasoline" = "Gasoline")

# Small-multiple overrides applied to every panel
sm <- theme(
  plot.title      = element_text(size = 10, face = "bold", margin = margin(b = 3)),
  axis.title      = element_text(size = 8),
  axis.text       = element_text(size = 7),
  legend.position = "none",
  plot.caption    = element_blank()
)

# A — Annual trend
pa <- ggplot(annual, aes(year, avg_price, colour = fuel)) +
  geom_vline(xintercept = c(2008, 2020, 2022),
             linetype = "dashed", colour = "grey82", linewidth = 0.4) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.8) +
  annotate("text", x = 2007.6, y = 4.15, label = "2008 crisis",
           size = 2.5, colour = "grey50", hjust = 1) +
  annotate("text", x = 2019.6, y = 0.95, label = "COVID",
           size = 2.5, colour = "grey50", hjust = 1) +
  annotate("text", x = 2021.6, y = 4.6, label = "Ukraine
war",
           size = 2.5, colour = "grey50", hjust = 1, lineheight = 0.85) +
  scale_colour_manual(values = fp, labels = fl, name = NULL) +
  scale_x_continuous(breaks = seq(1995, 2025, 5)) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Annual average prices", x = NULL, y = "USD / gallon") +
  sm

# B — Seasonal boxplots; fill derived from colour so guides merge to one legend
pb <- ggplot(prices, aes(month, price, colour = fuel,
                          fill = after_scale(alpha(colour, 0.25)))) +
  geom_boxplot(linewidth = 0.45, outlier.shape = NA) +
  scale_colour_manual(values = fp, labels = fl, name = NULL) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Seasonal distribution", x = NULL, y = "USD / gallon") +
  sm

# C — Week-on-week change
pc <- ggplot(diffs, aes(date, delta, colour = fuel)) +
  geom_hline(yintercept = 0, colour = "grey60", linewidth = 0.5) +
  geom_line(alpha = 0.55, linewidth = 0.35) +
  scale_colour_manual(values = fp, labels = fl, name = NULL) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Week-on-week price change", x = NULL, y = "Δ USD / gallon") +
  sm

# D — Scatter; trend line in diesel purple, no fuel colour scale
pd <- ggplot(wide, aes(gasoline, diesel)) +
  geom_point(alpha = 0.3, size = 0.9, colour = "grey55") +
  geom_smooth(method = "lm", colour = "#440154", linewidth = 0.9, se = FALSE) +
  annotate("text", x = Inf, y = -Inf, label = paste0("r = ", cor_val),
           size = 3, hjust = 1.1, vjust = -0.6, colour = "grey35",
           fontface = "italic") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = dollar_format()) +
  labs(title = "Diesel vs. gasoline", x = "Gasoline (USD / gal)", y = "Diesel (USD / gal)") +
  sm

# Assemble — single shared legend; all fuel aesthetics use scale_colour_manual
(pa + pb) / (pc + pd) +
  plot_layout(guides = "collect") +
  plot_annotation(
    title    = "Weekly US Fuel Prices: Trends, Seasonality & Volatility (1995–2025)",
    subtitle = "Diesel vs. Regular Conventional Gasoline · EIA via TidyTuesday",
    caption  = "Analysis: Joseph Nyajuoga"
  ) &
  theme(
    legend.position = "bottom",
    legend.key.size = unit(0.55, "cm"),
    legend.text     = element_text(size = 10),
    plot.title      = element_text(size = 15, face = "bold"),
    plot.subtitle   = element_text(size = 11, colour = "grey40"),
    plot.caption    = element_text(size = 9,  colour = "grey50")
  )
```

------------------------------------------------------------------------

## Conclusions

Three decades of weekly data point to a few durable findings:

- **Prices are structurally higher than they were in the 1990s.** Nominal prices have roughly tripled since 1995, with most of the step-change occurring during 2003-2008 as global demand outpaced refining capacity.
- **External shocks dominate the variance.** The 2008 financial crisis, 2020 pandemic, and 2022 Russia-Ukraine war each produced moves that dwarf the typical seasonal signal, making fuel prices a useful macro-economic indicator in their own right.
- **Diesel carries a persistent premium.** Before 2005, diesel was often cheaper than gasoline; since then it has traded at a consistent premium of roughly 20-50 cents per gallon, driven by global distillate demand from emerging markets and the heating-oil market.
- **Seasonality is real but modest.** Within any given year, price variation is roughly 10-20 cents, far smaller than the year-to-year shifts driven by crude oil prices.

```{r save-plots}
#| include: false
ggsave("images/annual_avg_prices.png",    p1,       width = 9,  height = 5, dpi = 300)
ggsave("images/monthly_distribution.png", p2,       width = 8,  height = 5, dpi = 300)
ggsave("images/weekly_change.png",        p3,       width = 9,  height = 5, dpi = 300)
ggsave("images/diesel_vs_gasoline.png",   p4,       width = 7,  height = 5, dpi = 300)
```